# load libraries and data

download worldclim (updated to use terra and sf)
https://gis.stackexchange.com/questions/227585/using-r-to-extract-data-from-worldclim
https://damariszurell.github.io/EEC-MGC/b2_EnvData.html
https://groups.google.com/g/MAXENT/c/9NWuqaqbO10

citation for grasshopper occurences:
GBIF.org (27 July 2023) GBIF Occurrence Download  https://doi.org/10.15468/dl.dqxb4h
```{r}
library(terra)
library(sf)
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(geodata)
library(dismo)
library(rJava) # note: need to install an x86_64 build of Java for this to work
# brew install --cask temurin
# R CMD javareconf
# install.packages("rJava", type = "source")
# Sys.setenv(JAVA_HOME = "/Library/Java/JavaVirtualMachines/temurin-17.jdk/Contents/Home")

# export JAVA_HOME=$(/usr/libexec/java_home -v 17)
              # sudo -E R CMD javareconf
sin = "+proj=sinu +lon_0=0 +x_0=0 +y_0=0 +R=6371007.181 +units=m +no_defs"
maxent()
outpath <- "/Users/mikea/Documents/mikedata/grasshopper/202502/"

# grasshopper
data_sources <- read.csv("data/datasets_download_usage_0118912-230530130749713.tsv", sep = "\t")
g <- read.csv("data/0118912-230530130749713.csv", sep = "\t") %>%
  rename(lat = decimalLatitude, lon = decimalLongitude) %>%
  mutate(dupcheck = paste0(lat,lon),
         dup = duplicated(dupcheck)) %>%
  filter(dup == FALSE) %>% 
  dplyr::select(-dupcheck, -dup)

# set path to worldclim tifs
wc_path <- "/Users/mikea/Documents/mikedata/cpm/202406/wc2"
wc_files <- list.files(wc_path, pattern = "*.tif", full.names = T)

# Read the GeoTIFF files into a raster stack
wc <- rast(wc_files)

plot(wc[[3]])

plot(wc[[1]])

env <- wc

# load grasshopper point locations
# remove 3 observations that were extreme western outliers (at least one of which was a disputed ID)
points <- g %>%
  filter(lon > -109) %>%
  dplyr::select(id = gbifID, lat, lon) %>%
  st_as_sf(., coords = c("lon", "lat"), crs = 4326)

# plot points
plot(env[[1]], xlim = c(-120, -50), ylim = c(15, 50))
plot(points, add = TRUE)

# this creates a 400 km
# that equates to about a 4-decimal-degree buffer per https://www.johndcook.com/how_big_is_a_degree.html
# occurrence data
occ_buff <- points %>%
  # st_transform(crs = sin) %>%
  st_buffer(., dist = 400000) %>%
  st_union()

# plot the first element ([[1]]) in the raster stack
plot(env[[1]], xlim = c(-109, -50), ylim = c(15, 50))
plot(occ_buff, add = T, col = "blue")  # adds buffer polygon to the plot
plot(points, add = T, col = "red")  # adds occurrence data to the plot

# crop study area to a manageable extent (rectangle shaped)
studyArea <- terra::crop(env, ext(as(occ_buff, "Spatial")))  

# the 'study area' created by extracting the buffer area from the raster stack
studyArea <- mask(studyArea, vect(occ_buff))

# output will still be a raster stack, just of the study area
plot(studyArea[[1]])
writeRaster(studyArea, paste0(outpath, "studyArea.tif"))

# # save the new study area rasters as ascii
# writeRaster(studyArea,
#             # a series of names for output files
#             filename=paste0(
#               "/Users/mikea/Documents/mikedata/grasshopper/202407/env_final/",
#                             names(studyArea),".wclc.asc"), 
#             # format="ascii", ## the output format
#             # bylayer=TRUE, ## this will save a series of layers
#             overwrite=T)

# select background points from this buffered area; when the number provided 
# to set.seed() function, the same random sample will be selected in the next line			
# use this code before the spatSample function every time, if you want to get
# the same "random samples"

# select background points from this buffered area
numbg <- 10000

set.seed(2); bg1 <- spatSample(x = studyArea,
                 size = numbg,
                 method = "random",
                 as.points = TRUE,
                 na.rm = TRUE) %>%
  as(., "Spatial") %>%
  st_as_sf(crs = 4326) %>%
  mutate(id = as.numeric(paste0(9999, 1:10000))) %>%
  dplyr::select(id)

set.seed(74); bg2 <- spatSample(x = studyArea,
                                size = numbg,
                                method = "random",
                                as.points = TRUE,
                                na.rm = TRUE) %>%
  as(., "Spatial") %>%
  st_as_sf(crs = 4326) %>%
  mutate(id = as.numeric(paste0(9999, 1:numbg))) %>%
  dplyr::select(id)

set.seed(94); bg3 <- spatSample(x = studyArea,
                                size = numbg,
                                method = "random",
                                as.points = TRUE,
                                na.rm = TRUE) %>%
  as(., "Spatial") %>%
  st_as_sf(crs = 4326) %>%
  mutate(id = as.numeric(paste0(9999, 1:numbg))) %>%
  dplyr::select(id)

set.seed(114); bg4 <- spatSample(x = studyArea,
                                size = numbg,
                                method = "random",
                                as.points = TRUE,
                                na.rm = TRUE) %>%
  as(., "Spatial") %>%
  st_as_sf(crs = 4326) %>%
  mutate(id = as.numeric(paste0(9999, 1:numbg))) %>%
  dplyr::select(id)

set.seed(134); bg5 <- spatSample(x = studyArea,
                                size = numbg,
                                method = "random",
                                as.points = TRUE,
                                na.rm = TRUE) %>%
  as(., "Spatial") %>%
  st_as_sf(crs = 4326) %>%
  mutate(id = as.numeric(paste0(9999, 1:numbg))) %>%
  dplyr::select(id)

# add the background & occurrence points to the plotted raster
plot(studyArea[[1]])
plot(bg1,add=T) 
plot(points, add=T, col="red")

# get environmental values from background points for modeling
bg1b <- cbind(bg1, terra::extract(env, bg1)) %>% dplyr::select(-ID)
bg2b <- cbind(bg1, terra::extract(env, bg2)) %>% dplyr::select(-ID)
bg3b <- cbind(bg1, terra::extract(env, bg3)) %>% dplyr::select(-ID)
bg4b <- cbind(bg1, terra::extract(env, bg4)) %>% dplyr::select(-ID)
bg5b <- cbind(bg1, terra::extract(env, bg5)) %>% dplyr::select(-ID)

# make a final point data set with observations (will be subsampled later)
train.points_sp <- cbind(points, terra::extract(env, points)) %>% dplyr::select(-ID) %>%
  as(., "Spatial") 

# make a final point data set with background points
bg.train.points1 <- bg1b %>%
  as.data.frame()

bg.train.points2 <- bg2b %>%
  as.data.frame()

bg.train.points3 <- bg3b %>%
  as.data.frame()

bg.train.points4 <- bg4b %>%
  as.data.frame()

bg.train.points5 <- bg5b %>%
  as.data.frame()

# spatially thin occurrence data (keep one occurrence point per cell)
cells <- terra::cellFromXY(raster(env[[16]]), train.points_sp)
dups <- duplicated(cells)
occ_final <- train.points_sp[!dups, ] %>%
  st_as_sf(., crs = st_crs(env))
cat(nrow(train.points_sp) - nrow(occ_final), "records are removed")
st_crs(occ_final) <- st_crs(env)


```
# Build maxent model using dismo
build a maxent model in dismo
https://rdrr.io/cran/dismo/man/maxent.html
https://github.com/shandongfx/workshop_maxent_R/blob/master/code/Appendix1_case_study.md
Note: to make maxent work, I had to download a version of R that matched the build of Java that was installed. 
```{r}

maxent()

### Build maxent model using dismo
# https://rdrr.io/cran/dismo/man/maxent.html
# https://github.com/shandongfx/workshop_maxent_R/blob/master/code/Appendix1_case_study.md

# get the random samples of occurrence points for training and testing

# randomly select 70% for training (5 sets)
set.seed(2); selected1 <- sample(1:nrow(occ_final), nrow(occ_final) * 0.7, replace = F)
set.seed(20); selected2 <- sample(1:nrow(occ_final), nrow(occ_final) * 0.7, replace = F)
set.seed(200); selected3 <- sample(1:nrow(occ_final), nrow(occ_final) * 0.7, replace = F)
set.seed(2000); selected4 <- sample(1:nrow(occ_final), nrow(occ_final) * 0.7, replace = F)
set.seed(2100); selected5 <- sample(1:nrow(occ_final), nrow(occ_final) * 0.7, replace = F)

occ_train1 <- occ_final[selected1, ]  # this is the selection to be used for model training
occ_test1 <- occ_final[-selected1, ]  # this is the opposite of the selection which will be used for model testing
occ_train2 <- occ_final[selected2, ]  # this is the selection to be used for model training
occ_test2 <- occ_final[-selected2, ]  # this is the opposite of the selection which will be used for model testing
occ_train3 <- occ_final[selected3, ]  # this is the selection to be used for model training
occ_test3 <- occ_final[-selected3, ]  # this is the opposite of the selection which will be used for model testing
occ_train4 <- occ_final[selected4, ]  # this is the selection to be used for model training
occ_test4 <- occ_final[-selected4, ]  # this is the opposite of the selection which will be used for model testing
occ_train5 <- occ_final[selected5, ]  # this is the selection to be used for model training
occ_test5 <- occ_final[-selected5, ]  # this is the opposite of the selection which will be used for model testing

# env conditions for training occurrences from the raster
p.env1 <- occ_train1 %>%
  st_drop_geometry() %>%
  as.data.frame() %>%
  dplyr::select(-id) %>%
  mutate(pa = 1) %>%
  bind_rows(st_drop_geometry(mutate(bg.train.points1, pa = 0))) %>%
  dplyr::select(-id, -geometry)

p.env2 <- occ_train2 %>%
  st_drop_geometry() %>%
  as.data.frame() %>%
  dplyr::select(-id) %>%
  mutate(pa = 1) %>%
  bind_rows(st_drop_geometry(mutate(bg.train.points2, pa = 0))) %>%
  dplyr::select(-id, -geometry)

p.env3 <- occ_train3 %>%
  st_drop_geometry() %>%
  as.data.frame() %>%
  dplyr::select(-id) %>%
  mutate(pa = 1) %>%
  bind_rows(st_drop_geometry(mutate(bg.train.points3, pa = 0))) %>%
  dplyr::select(-id, -geometry)

p.env4 <- occ_train4 %>%
  st_drop_geometry() %>%
  as.data.frame() %>%
  dplyr::select(-id) %>%
  mutate(pa = 1) %>%
  bind_rows(st_drop_geometry(mutate(bg.train.points4, pa = 0))) %>%
  dplyr::select(-id, -geometry)

p.env5 <- occ_train5 %>%
  st_drop_geometry() %>%
  as.data.frame() %>%
  dplyr::select(-id) %>%
  mutate(pa = 1) %>%
  bind_rows(st_drop_geometry(mutate(bg.train.points5, pa = 0))) %>%
  dplyr::select(-id, -geometry)

# just the presence observations
p.env1.pres <- p.env1 %>%
  filter(pa %in% 1) %>%
  dplyr::select(-pa)

p.env2.pres <- p.env2 %>%
  filter(pa %in% 1) %>%
  dplyr::select(-pa)

p.env3.pres <- p.env3 %>%
  filter(pa %in% 1) %>%
  dplyr::select(-pa)

p.env4.pres <- p.env4 %>%
  filter(pa %in% 1) %>%
  dplyr::select(-pa)

p.env5.pres <- p.env5 %>%
  filter(pa %in% 1) %>%
  dplyr::select(-pa)

# env conditions for testing occurrence (presence) points
p.env1_test.pres <- occ_test1 %>%
  st_drop_geometry() %>%
  as.data.frame() %>%
  dplyr::select(-id)

p.env2_test.pres <- occ_test2 %>%
  st_drop_geometry() %>%
  as.data.frame() %>%
  dplyr::select(-id)

p.env3_test.pres <- occ_test3 %>%
  st_drop_geometry() %>%
  as.data.frame() %>%
  dplyr::select(-id)

p.env4_test.pres <- occ_test4 %>%
  st_drop_geometry() %>%
  as.data.frame() %>%
  dplyr::select(-id)

p.env5_test.pres <- occ_test5 %>%
  st_drop_geometry() %>%
  as.data.frame() %>%
  dplyr::select(-id)


# train Maxent with tabular data
mod1 <- maxent(x=dplyr::select(p.env1, -pa), ## env conditions
              p=p.env1$pa,   ## 1:presence or 0:absence
              path="/Users/mikea/Documents/research/grasshopper/output/maxent_outputs_no_lc1", 
              args=c("responsecurves") ## parameter specification
)

mod2 <- maxent(x=dplyr::select(p.env2, -pa), ## env conditions
               p=p.env2$pa,   ## 1:presence or 0:absence
              path="/Users/mikea/Documents/research/grasshopper/output/maxent_outputs_no_lc2", 
               args=c("responsecurves") ## parameter specification
)

mod3 <- maxent(x=dplyr::select(p.env3, -pa), ## env conditions
               p=p.env3$pa,   ## 1:presence or 0:absence
              path="/Users/mikea/Documents/research/grasshopper/output/maxent_outputs_no_lc3", 
               args=c("responsecurves") ## parameter specification
)

mod4 <- maxent(x=dplyr::select(p.env4, -pa), ## env conditions
               p=p.env4$pa,   ## 1:presence or 0:absence
              path="/Users/mikea/Documents/research/grasshopper/output/maxent_outputs_no_lc4", 
               args=c("responsecurves") ## parameter specification
)

mod5 <- maxent(x=dplyr::select(p.env5, -pa), ## env conditions
               p=p.env5$pa,   ## 1:presence or 0:absence
              path="/Users/mikea/Documents/research/grasshopper/output/maxent_outputs_no_lc5", 
               args=c("responsecurves") ## parameter specification
)

# save model output
saveRDS(mod1, paste0(outpath, "mod1_grasshopper.rds"))
saveRDS(mod2, paste0(outpath, "mod2_grasshopper.rds"))
saveRDS(mod3, paste0(outpath, "mod3_grasshopper.rds"))
saveRDS(mod4, paste0(outpath, "mod4_grasshopper.rds"))
saveRDS(mod5, paste0(outpath, "mod5_grasshopper.rds"))

# view the maxent models in a html brower
# show(mod1)
# show(mod2)
# show(mod3)
# show(mod4)
# show(mod5)

env_crop <- env %>%
  crop(studyArea)

# make predictions for Study Area 
ped1 <- predict(mod1,raster::stack(studyArea))
ped2 <- predict(mod2,raster::stack(studyArea))
ped3 <- predict(mod3,raster::stack(studyArea))
ped4 <- predict(mod4,raster::stack(studyArea))
ped5 <- predict(mod5,raster::stack(studyArea))

# make predictions for training occurrences
tped1 <- predict(mod1, p.env1.pres); (tpq1 <- quantile(tped1, c(0.1))) # .3752
tped2 <- predict(mod2, p.env2.pres); (tpq2 <- quantile(tped2, c(0.1))) # 
tped3 <- predict(mod3, p.env3.pres); (tpq3 <- quantile(tped3, c(0.1))) # 
tped4 <- predict(mod4, p.env4.pres); (tpq4 <- quantile(tped4, c(0.1))) # 
tped5 <- predict(mod5, p.env5.pres); (tpq5 <- quantile(tped5, c(0.1))) # 

# using 'training data' to evaluate model performance
(mod1_eval_train <- dismo::evaluate(p = p.env1.pres, a = bg.train.points1, model = mod1)) # 0.91
(mod2_eval_train <- dismo::evaluate(p = p.env2.pres, a = bg.train.points2, model = mod2))
(mod3_eval_train <- dismo::evaluate(p = p.env3.pres, a = bg.train.points3, model = mod3))
(mod4_eval_train <- dismo::evaluate(p = p.env4.pres, a = bg.train.points4, model = mod4))
(mod5_eval_train <- dismo::evaluate(p = p.env5.pres, a = bg.train.points5, model = mod5))

# get test evaluation stats
(mod1_eval_test <- dismo::evaluate(p = p.env1_test.pres, a = bg.train.points1, model = mod1)) # 0.908
(mod2_eval_test <- dismo::evaluate(p = p.env2_test.pres, a = bg.train.points2, model = mod2))
(mod3_eval_test <- dismo::evaluate(p = p.env3_test.pres, a = bg.train.points3, model = mod3))
(mod4_eval_test <- dismo::evaluate(p = p.env4_test.pres, a = bg.train.points4, model = mod4))
(mod5_eval_test <- dismo::evaluate(p = p.env5_test.pres, a = bg.train.points5, model = mod5))

# calculate thresholds of models
# (thd1 <- threshold(mod_eval_train, "no_omission"))  # 0% omission rate
# (thd2 <- threshold(mod_eval_train, "spec_sens"))  # highest TSS

# write rasters for individual model runs
ped_all <- stack(ped1, ped2, ped3, ped4, ped5)

mean_pred <- mean(ped_all)

# plotting points that are above the previously calculated
# thresholded value
mean(tpq1, tpq2, tpq3, tpq4, tpq5) # 0.375
plot(mean_pred > .375)
plot(occ_final, add = T)
mean_pred10pct <- mean_pred
mean_pred10pct <- (mean_pred10pct > 0.375)
plot(mean_pred10pct)


# write rasters
writeRaster(ped_all, paste0(outpath, "grasshopper_present_5mods.tif"), overwrite = T)
writeRaster(mean_pred, paste0(outpath, "grasshopper_present_5mod_avg.tif"), overwrite = T)
writeRaster(mean_pred10pct, paste0(outpath, "grasshopper_present_5mod_avg_10pct.tif"), overwrite = T)


png("figures/grasshopper_present_5mod_avg.png", width = 5*600, height = 5*600, res = 600)
plot(mean_pred)
dev.off()
plot(mean_pred)

# collect model stats
mod_AUC <- data.frame(model = c("mod1", "mod2", "mod3", "mod4", "mod5", "mean"),
                      vars = "all",
                      trainAUC = c(mod1_eval_train@auc, mod2_eval_train@auc,
                                   mod3_eval_train@auc, mod4_eval_train@auc,
                                   mod5_eval_train@auc, NaN),
                      testAUC = c(mod1_eval_test@auc, mod2_eval_test@auc,
                                  mod3_eval_test@auc, mod4_eval_test@auc,
                                  mod5_eval_test@auc, NaN))
mod_AUC[6,3] <- mean(mod_AUC$trainAUC, na.rm = T)
mod_AUC[6,4] <- mean(mod_AUC$testAUC, na.rm = T)
mod_AUC

write.csv(mod_AUC, "output/mod_AUC_grasshopper.csv")



```
# future climate data
https://www.worldclim.org/data/cmip6/cmip6_clim5m.html
```{r}

# Download future climate scenario from 'ACCESS-ESM1-5' climate model.
# Please note that you have to set download=T if you haven't downloaded the data before:
clim_fut <- geodata::cmip6_world(model='ACCESS-ESM1-5', ssp='245', time='2061-2080', var='bioc', download=F, res=2.5, path="/Users/mikea/Documents/mikedata/grasshopper/202407/")

# the future 'study area' created by extracting the buffer area from the raster stack
studyArea.future <- clim_fut %>%
  crop(., studyArea) %>%
  mask(., studyArea)
(namevect <- names(studyArea.future))
names(studyArea)

names(studyArea.future) <- paste0(substr(namevect,1,14), "_", substr(namevect,48,49))

ped1.future <- predict(mod1, raster::stack(studyArea.future))  
ped2.future <- predict(mod2, raster::stack(studyArea.future))  
ped3.future <- predict(mod3, raster::stack(studyArea.future))  
ped4.future <- predict(mod4, raster::stack(studyArea.future))  
ped5.future <- predict(mod5, raster::stack(studyArea.future))  

plot(mean_pred)
plot(ped1.future)  # plot the continuous prediction

# combine individual model runs
ped_all.future <- stack(ped1.future, ped2.future, 
                        ped3.future, ped4.future, 
                        ped5.future)

mean_pred.future <- mean(ped_all.future)
plot(mean_pred.future)
writeRaster(mean_pred.future, paste0(outpath, "grasshopper_present_5mod_avg.future.tif"), overwrite = T)



```


# current & future plots (North America)
```{r}
mean_pred <- terra::rast("/Users/mikea/Documents/mikedata/grasshopper/202502/grasshopper_present_5mod_avg.tif")
studyArea <- terra::rast("/Users/mikea/Documents/mikedata/grasshopper/202502/studyArea.tif")
mod1 <- readRDS(paste0(outpath, "mod1_grasshopper.rds"))
mod2 <- readRDS(paste0(outpath, "mod2_grasshopper.rds"))
mod3 <- readRDS(paste0(outpath, "mod3_grasshopper.rds"))
mod4 <- readRDS(paste0(outpath, "mod4_grasshopper.rds"))
mod5 <- readRDS(paste0(outpath, "mod5_grasshopper.rds"))
mean_pred.future <- rast(paste0(outpath, "grasshopper_present_5mod_avg.future.tif"))
# us <- rnaturalearthhires::countries10
world_highres <- rnaturalearth::ne_countries(scale = 10, returnclass = "sf")
lakes <- rnaturalearth::ne_download(scale = 110, type = "lakes", category = "physical", returnclass = "sf")


mean_pred10pct.future <- mean_pred.future
mean_pred10pct.future <- (mean_pred10pct.future > 0.375)
# mean_pred10pct.future <- rast(mean_pred10pct.future)
plot(mean_pred10pct.future)
mean_pred10pct <- mean_pred
mean_pred10pct <- (mean_pred10pct > 0.375)*1
plot(mean_pred10pct)

writeRaster(mean_pred10pct.future, paste0(outpath, "grasshopper_present_5mod_avg_10pct.future.tif"), overwrite = T)

current_df <- as.data.frame(mean_pred, xy = TRUE) %>%
  dplyr::rename(value = 3) %>%
  filter(value > 0.375)
future_df <- as.data.frame(mean_pred.future, xy = TRUE) %>%
  dplyr::rename(value = 3) %>%
  filter(value > 0.375)

current_df.bin <- as.data.frame(mean_pred10pct, xy = TRUE) %>%
  dplyr::rename(value = 3) %>%
  filter(value %in% 1)
future_df.bin <- as.data.frame(mean_pred10pct.future, xy = TRUE) %>%
  dplyr::rename(value = 3) %>%
  filter(value %in% 1)

(cur.bin <- ggplot(data = current_df.bin) +
  geom_sf(data = world_highres) +
  geom_tile(aes(x = x, y = y, fill = value), fill = "darkred") + 
  geom_sf(data = lakes, fill = "steelblue") +
  lims(x = c(-104, -68), y = c(20,50)) +
  theme_minimal() +
  labs(x = "", y = "", fill = "", title = "1970-2000") +
  guides(fill = "none"))

(fut.bin <- ggplot(data = future_df.bin) +
  geom_sf(data = world_highres) +
  geom_tile(aes(x = x, y = y, fill = value), fill = "darkred") + 
  geom_sf(data = lakes, fill = "steelblue") +
  lims(x = c(-104, -68), y = c(20,50)) +
  theme_minimal() +
  labs(x = "", y = "", fill = "", title = "2061-2080") +
  guides(fill = "none"))

library(patchwork)
cur.bin + fut.bin

ggsave("figures/grasshopper_current_future_5mod_avg.bin2.png", height = 4, width = 8, dpi = 600)

(cur <- ggplot(data = current_df) +
  geom_sf(data = world_highres) +
  geom_tile(aes(x = x, y = y, fill = value)) + 
  scale_fill_viridis_c(option = "inferno") +
  geom_sf(data = lakes, fill = "steelblue") +
  lims(x = c(-104, -68), y = c(20,50)) +
  theme_minimal() +
  labs(x = "", y = "", fill = "") + # , title = "1970-2000"
  guides(fill = "none") +
    theme(axis.text = element_blank()))

(fut <- ggplot(data = future_df) +
  geom_sf(data = world_highres) +
  geom_tile(aes(x = x, y = y, fill = value)) + 
  scale_fill_viridis_c(option = "inferno") +
  geom_sf(data = lakes, fill = "steelblue") +
  lims(x = c(-104, -68), y = c(20,50)) +
  theme_minimal() +
  labs(x = "", y = "", fill = "") + # , title = "2061-2080"
  guides(fill = "none") +
    theme(axis.text = element_blank()))

library(patchwork)
# cur + fut

# ggsave("figures/grasshopper_current_future_5mod_avg.viridis.png", height = 4, width = 8, dpi = 600)

(both <- ggplot(data = future_df.bin) +
  geom_sf(data = world_highres) +
  geom_tile(aes(x = x, y = y), 
            fill = "darkred",
            alpha = 0.5) +
  geom_tile(aes(x = x, y = y), 
            fill = "blue", 
            alpha = 0.5,
            data = current_df.bin) +
  geom_sf(data = lakes, fill = "steelblue") +
  lims(x = c(-104, -68), y = c(20,50)) +
  theme_minimal() +
  labs(x = "", y = "") +
  guides(fill = "none"))

both + cur/fut + plot_layout(widths = c(2.5, 1))

ggsave("figures/grasshopper_current_future_5mod_avg.bin.3panel.png", height = 8, width = 8, dpi = 600, bg = "white")

# current range size
(num_cells_1 <- sum(current_df.bin$value %in% 1))
cur.bin.proj = terra::project(mean_pred10pct, sin)
cell_area <- prod(res(cur.bin.proj))
# Calculate total area
(area_1 <- num_cells_1 * cell_area / 1000000)# 765566.2 km2

# future range size
(num_cells_1.fut <- sum(future_df.bin$value %in% 1))
fut.bin.proj = terra::project(mean_pred10pct.future, sin)
cell_area.fut <- prod(res(fut.bin.proj))
# Calculate total area
(area_1.fut <- num_cells_1.fut * cell_area.fut / 1000000)# 765566.2 km2

```
# model performance
```{r}


modAUC <- read.csv("output/mod_AUC_grasshopper.csv")
summary(modAUC$trainAUC[1:5])
summary(modAUC$testAUC[1:5])
```
# plot incursions
```{r}

ir <- g %>%
  mutate(n.sightings = case_when(lat > 38 ~ 1,
                                TRUE ~ 0),
         n.state = case_when(stateProvince %in% c("New Jersey",
                                                 "New York",
                                                 "Pennsylvania") ~ 1,
                            TRUE ~ 0)) %>%
  group_by(year) %>%
  summarize(n.sightings = sum(n.sightings),
            n.state = sum(n.state),
            n = length(year),
            .groups = "drop") %>%
  mutate(pct = 100*(n.sightings/n),
         pct.state = 100*(n.state/n))

ir %>%
  filter(year > 1990,
         n > 20) %>%
ggplot() +
  geom_line(aes(x = year, y = pct)) +
  geom_point(aes(x = year, y = pct)) +
  geom_text(aes(x = year, y = pct+1, label = n), color = "darkgray") +
  scale_x_continuous(breaks = seq(2013, 2023, by = 2)) +
  theme_bw() +
  theme(text = element_text(size = 14)) +
  labs(x = "", y = "% of sightings\nabove 38 N latitude")

ggsave("figures/grasshopper_pct_sightings_above_38N.png", height = 4, width = 4, dpi = 600, bg = "white")

```


# predict present and future climate data (entire world)
https://www.worldclim.org/data/cmip6/cmip6_clim5m.html
```{r}

# Download future climate scenario from 'ACCESS-ESM1-5' climate model.
# Please note that you have to set download=T if you haven't downloaded the data before:
clim_fut <- geodata::cmip6_world(model='ACCESS-ESM1-5', ssp='245', time='2061-2080', var='bioc', download=F, res=2.5, path="/Users/mikea/Documents/mikedata/grasshopper/202407/")

# the future 'study area' created by extracting the buffer area from the raster stack
studyArea.future <- clim_fut %>%
  crop(., studyArea) %>%
  mask(., studyArea)
(namevect <- names(studyArea.future))
names(studyArea)

names(clim_fut) <- paste0(substr(namevect,1,14), "_", substr(namevect,48,49))

ped1.world <- predict(mod1, raster::stack(wc))  
ped2.world <- predict(mod2, raster::stack(wc))  
ped3.world <- predict(mod3, raster::stack(wc))  
ped4.world <- predict(mod4, raster::stack(wc))  
ped5.world <- predict(mod5, raster::stack(wc))  
ped1.world.future <- predict(mod1, raster::stack(clim_fut))  
ped2.world.future <- predict(mod2, raster::stack(clim_fut))  
ped3.world.future <- predict(mod3, raster::stack(clim_fut))  
ped4.world.future <- predict(mod4, raster::stack(clim_fut))  
ped5.world.future <- predict(mod5, raster::stack(clim_fut))  

# combine individual model runs
ped_all.world <- stack(ped1.world, ped2.world, 
                        ped3.world, ped4.world, 
                        ped5.world)
ped_all.world.future <- stack(ped1.world.future, ped2.world.future, 
                        ped3.world.future, ped4.world.future, 
                        ped5.world.future)

mean_pred.world <- mean(ped_all.world)
mean_pred.world.future <- mean(ped_all.world.future)

# write the raster to big data folder
writeRaster(mean_pred.world, paste0(outpath, "grasshopper_present_5mod_avg.world.tif"), overwrite = T)

writeRaster(mean_pred.world.future, paste0(outpath, "grasshopper_present_5mod_avg.world.future.tif"), overwrite = T)



```


# current & future plots (entire world)
```{r}
mean_pred.world <- terra::rast("/Users/mikea/Documents/mikedata/grasshopper/202502/grasshopper_present_5mod_avg.world.tif")

mean_pred.world.future <- terra::rast("/Users/mikea/Documents/mikedata/grasshopper/202502/grasshopper_present_5mod_avg.world.future.tif")

# us <- rnaturalearthhires::countries10
world_highres <- rnaturalearth::ne_countries(scale = 10, returnclass = "sf")
lakes <- rnaturalearth::ne_download(scale = 110, type = "lakes", category = "physical", returnclass = "sf")

# Reproject vector data to Mollweide
world_highres_moll <- st_transform(world_highres, crs = 'ESRI:54009')
lakes_moll <- st_transform(lakes, crs = 'ESRI:54009')

# get 90% distribution (world, future and current)
mean_pred10pct.world.future <- mean_pred.world.future
mean_pred10pct.world.future <- (mean_pred10pct.world.future > 0.375)

mean_pred10pct.world <- mean_pred.world
mean_pred10pct.world <- (mean_pred10pct.world > 0.375)*1

# Reproject the rasters into Mollweide
mean_pred10pct.world_moll <- 
  terra::project(rast(mean_pred10pct.world), 
                                            "ESRI:54009")
mean_pred10pct.world.future_moll <- 
  terra::project(rast(mean_pred10pct.world.future), 
                                                   "ESRI:54009")
mean_pred.world_moll <- 
  terra::project(rast(mean_pred.world), 
                                            "ESRI:54009")
mean_pred.world.future_moll <- 
  terra::project(rast(mean_pred.world.future), 
                                                   "ESRI:54009")

writeRaster(mean_pred10pct.world_moll, paste0(outpath, "grasshopper_present_5mod_avg_10pct.world_moll.tif"), overwrite = T)
writeRaster(mean_pred10pct.world.future_moll, paste0(outpath, "grasshopper_present_5mod_avg_10pct.world.future_moll.tif"), overwrite = T)
writeRaster(mean_pred.world_moll, paste0(outpath, "grasshopper_present_5mod_avg.world_moll.tif"), overwrite = T)
writeRaster(mean_pred.world.future_moll, paste0(outpath, "grasshopper_present_5mod_avg.world.future_moll.tif"), overwrite = T)

# make dataframe of current and future continous predictions for ggplot (world)
current_df.world <- as.data.frame(mean_pred.world_moll, xy = TRUE) %>%
  dplyr::rename(value = 3) %>%
  filter(value > 0.375)
future_df.world <- as.data.frame(mean_pred.world.future_moll, xy = TRUE) %>%
  dplyr::rename(value = 3) %>%
  filter(value > 0.375)

# convert binary rasters to to dataframe for ggplot
current_df.world.bin <- as.data.frame(mean_pred10pct.world_moll, 
                                      xy = TRUE) %>%
  dplyr::rename(value = 3) %>%
  filter(value %in% 1)
future_df.world.bin <- as.data.frame(mean_pred10pct.world.future_moll, 
                                     xy = TRUE) %>%
  dplyr::rename(value = 3) %>%
  filter(value %in% 1)

# plot continuous predictions (entire world), current and future
# (cur <- ggplot(data = current_df.world) +
#   geom_sf(data = world_highres_moll) +
#   geom_tile(aes(x = x, y = y, fill = value)) +
#   scale_fill_viridis_c(option = "inferno") +
#   geom_sf(data = lakes_moll, fill = "steelblue") +
#   # lims(x = c(-104, -68), y = c(20,50)) +
#   coord_sf(crs = 'ESRI:54009') +  # Mollweide projection
#   theme_minimal() +
#   labs(x = "", y = "", fill = "") + # , title = "1970-2000"
#   guides(fill = "none") +
#     theme(axis.text = element_blank()))
# 
# (fut <- ggplot(data = future_df.world) +
#   geom_sf(data = world_highres_moll) +
#   geom_tile(aes(x = x, y = y, fill = value)) +
#   scale_fill_viridis_c(option = "inferno") +
#   geom_sf(data = lakes_moll, fill = "steelblue") +
#   # lims(x = c(-104, -68), y = c(20,50)) +  
#   coord_sf(crs = 'ESRI:54009') +  # Mollweide projection
#   theme_minimal() +
#   labs(x = "", y = "", fill = "") + # , title = "2061-2080"
#   guides(fill = "none") +
#     theme(axis.text = element_blank()))

(both <- ggplot(data = future_df.world.bin) +
  geom_sf(data = world_highres_moll, alpha = 0.5) +
  geom_tile(aes(x = x, y = y), 
            fill = "darkred",
            alpha = 0.5) +
  geom_tile(aes(x = x, y = y), 
            fill = "blue", 
            alpha = 0.5,
            data = current_df.world.bin) +
  geom_sf(data = lakes_moll, fill = "steelblue") +
  # lims(y = c(-60, 60)) +
  # lims(x = c(-104, -68), y = c(20,50)) +
    lims(y = c(-6184829-500, 6927685+900000)) +
  coord_sf(crs = 'ESRI:54009') +  # Mollweide projection
  theme_minimal() +
  labs(x = "", y = "") +
  guides(fill = "none"))

ggsave("figures/grasshopper_current_future_5mod_avg.world.bin3.png", height = 8, width = 8, dpi = 600, bg = "white")

# library(patchwork)
# both + cur/fut + plot_layout(widths = c(2.5, 1))
# 
# ggsave("figures/grasshopper_current_future_5mod_avg.world.bin.3panel.png", height = 8, width = 8, dpi = 600, bg = "white")

# current range size
(num_cells_1 <- sum(current_df.world.bin$value %in% 1))
cur.bin.proj = terra::project(rast(mean_pred10pct.world), sin)
cell_area <- prod(res(cur.bin.proj))
# Calculate total area
(area_1 <- num_cells_1 * cell_area / 1000000) # 7603502 km2

# future range size
(num_cells_1.fut <- sum(future_df.world.bin$value %in% 1))
fut.bin.proj = terra::project(rast(mean_pred10pct.world.future), sin)
cell_area.fut <- prod(res(fut.bin.proj))
# Calculate total area
(area_1.fut <- num_cells_1.fut * cell_area.fut / 1000000) # 9632759 km2

# calculate % increase in suitable area
(9632759 - 7603502) / 9632759
# [1] 0.2106621 # 21% increase in area
```